Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-24 paper.tex
updates until end of geometry section

\todo{LC: mention some results here and how they make $\rnni$ differ from $\nni$}
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
However, the Weak Split Theorem remains a conjecture.
Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).

@@

\begin{algorithm}[H]
\caption{$\mdtree(T)$}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease
	\FOR {taxon $a$ in $L$}
		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
	\ENDFOR
	\RETURN $R$
\end{algorithmic}
\end{algorithm}

@@

    \todo{LC: replace this figure}
    \begin{figure}[H]
    	\centering
    	\includegraphics[width=\textwidth]{cat_convex_seq_example}
    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ as described in proof of Theorem~\ref{thm:caterpillar_convex}.
        The trees are represented as sequences that are being ordered by $\csort$.
        Right before loop number $k$, $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
    	\label{fig:cat_convex_seq_example}
    \end{figure}

    Since it is clear that $T'$ has caterpillar distance $d_c(T',R) = l+1$ to $T$, we can apply the induction hypothesis on $T'$ and $R$ and see that $d(T',R) = l+1$.
    Furthermore, it is $d(T,T') = 1$.
    If it now were $d(T,R) < l$, it would hold $d(T',R) \leq d(T,R) + d(T,T') < l + 1$ which contradicts the induction hypothesis.
    We can conclude $d(T,R) = l$, which completes the proof.
\end{proof}

@@

With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
Furthermore, we are able to show that the \emph{radius} of the $\rnni$ graph, which is defined as $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ for a graph $G$, equals its diameter.
Let us first consider the diameter of the $\rnni$ graph.
With the results of Section~\ref{section:caterpillar_convex} is not hard to see that is $\Delta(\rnni) = \frac{(n-1)(n-2))}{2}$ as it is stated in Corollary~\ref{corollary:diameter}.

@@

\begin{proof}
	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
    We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
    When running this algorithm on $T$ and $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its right neighbour.
    Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.
\end{proof}

@@

We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
Therefore we use the algorithm $\mdtree$ that has been introduced in section~\ref{section:alg_mdtree}.

However, the results provided in this section are only true for caterpillar trees.
The pseudocode for $\mdtree$ can be found in Section~\ref{section:alg_mdtree}.
\todo{figure? + do we need the pseudocode?}

@@

\begin{proof}
    Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
    Within this proof we assume, without loss of generality, that it is $T = (( \dots (1,2)3)\dots)n)$
    We prove the lemma by induction on the number of taxa $n$.

    Induction basis: $n = 3$\\
    $\mdtree(T)$ results either in $R = ((1,3),2)$ or $R'= ((2,3),1)$ for $T = ((1,2),3)$.
    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa $\mdtree$ computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Let us now  assume that $T$ is a tree on $n+1$ taxa.
    In the list $L$ produced in $\mdtree(T)$ the taxon $n+1$ is the last one, $L[n] = n+1$.
    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
    It is easy to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$.
    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In $\mdtree$ it is not uniquely determined, where a new taxon is added to the already existing tree.
Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
Therefore the output tree $R$ of $\mdtree$ is not unique.
Every tree topology on $n$ taxa can be resulting from $\mdtree$.
And because permuting labels of trees does not change the distance between them, $\mdtree$ shows that there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
This gives us Corollary~\ref{corollary:radius}.

@@

\subsection{Split Theorem}
\label{section:split_theorem}
\todo{Split or Cluster?}

>>>>

\todo{LC: mention some results here and how they make $\rnni$ differ from $\nni$}
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
However, the Weak Split Theorem remains a conjecture.
Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:cluster_theorem}).

@@

\begin{algorithm}[H]
\caption{$\mdtree(T)$}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease \label{alg:mdtree_lineL}
	\FOR {taxon $a$ in $L$}
		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node \label{alg:mdtree_lineAddTaxon}
	\ENDFOR
	\RETURN $R$
\end{algorithmic}
\end{algorithm}

An important property of $\mdtree$, which we will use later on in Section~\ref{section:diameter} is the fact that this algorithm is non-deterministic.
This is because the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined.
Therefore the output tree $R$ of $\mdtree$ is not unique.

@@

    \todo{LC: replace this figure}
    \begin{figure}[H]
    	\centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
            \hline
            $x_{k+1}$ & $x_k$ & $\ldots$ & $b$ & $a$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $b$ & $x_k$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_k$ & $\ldots$ & $a$ & $b$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $x_k$ & $b$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
        \end{tabular}
        \vspace{2pt}
    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ (in list representation) as described in proof of Theorem~\ref{thm:caterpillar_convex}.
        The trees are represented as sequences that are being sorted by $\csort$.
        Right before loop number $k$, taxa $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
    	\label{fig:cat_convex_seq_example}
    \end{figure}

    Because $T'$ has caterpillar distance $d_c(T',R) = l+1$ to $T$, we can apply the induction hypothesis on $T'$ and $R$ and see that $d(T',R) = l+1$.
    Furthermore, it is $d(T,T') = 1$.
    If it now were $d(T,R) < l$, it would hold $d(T',R) \leq d(T,R) + d(T,T') < l + 1$ which contradicts the induction hypothesis.
    We can conclude $d(T,R) = l$, which completes the proof.
\end{proof}

@@

With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
Furthermore, we are able to show that the \emph{radius} of the $\rnni$ graph, which is defined as $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ for a graph $G$, equals its diameter.
At first we consider the diameter of the $\rnni$ graph.
By Lemma~\ref{lemma:diameter_bound} we already know an upper bound for the diameter.
We will now show that this is not only a bound, but the exact diameter of $\rnni$.

@@

\begin{proof}
	Let $T = [1,n,2,3,\ldots,n-1]$ and $R = [n-1,n,n-2,n-3, \ldtos, 1]$ be two caterpillar trees.
    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
    We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
    When running this algorithm on $T$ and $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its upper neighbour.
    Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.
\end{proof}

@@

We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
Therefore we use the algorithm $\mdtree$ that has been introduced in section~\ref{section:alg_mdtree}.
Note that the results provided in this section are only true if the input tree of $\mdtree$ is a caterpillar tree.

@@

\begin{proof}
    Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
    Within this proof we assume, without loss of generality, that it is $T = [1,2,3,\ldots,n]$.
    We prove the lemma by induction on the number of taxa $n$.

    Induction basis: $n = 3$\\
    $\mdtree(T)$ results either in $R =[1,3,2]$ or $R'= [2,3,1]$ for $T = [1,2,3]$.
    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa $\mdtree$ computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Let us now  assume that $T$ is a tree on $n+1$ taxa.
    In the list $L$ produced Line~\ref{alg:mdtree_lineL} in $\mdtree(T)$ (Algorithm~\ref{alg:max_dist_tree}) taxon $n+1$ is the last one, because its parent, the root, has the highest rank in $T$.
    This means that this taxon $n+1$ is the last one added last to $R$ as in Line~\ref{alg:mdtree_lineAddTaxon} and therefore, it has parent with rank one in $R$.
    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
    It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
    $R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alf:max_dist_tree} for $T|_n$ is the same as the one for $T$, just the last element is missing.
    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
In fact, every tree topology on $n$ taxa can be resulting from $\mdtree$, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{ald:max_dist_tree}.
This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
\todo{LC: Definition tree topology?}
And because permuting labels of trees does not change the distance between them, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
This gives us Corollary~\ref{corollary:radius}.

@@

\subsection{Cluster Theorem}
\label{section:cluster_theorem}
\todo{Split or Cluster?}
